---
title: "S7 in recast: objects on top of vectors"
vignette: >
  %\VignetteIndexEntry{S7 in recast: objects on top of vectors}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(recast)
```

[S7](https://rconsortium.github.io/S7/) is the R Consortium's next-generation
object system — a single, well-specified successor to S3 and S4, built by the
same working group that maintains R's object model, and intended one day to
land in base R. The tidyverse has been moving its *object* layer onto it. This
vignette is a field guide to adopting S7 in a package that already leans on
**vctrs** and **S3**, written from `recast`'s own migration: the `<mdl_gt>` table
specification and its resolved build were rewritten as S7 classes, and
every decision — including the ones that bit — is recorded here with the real
code.

The single most important idea comes first, because it shapes everything else.

## S7 does not replace vctrs — it sits above it

`recast` has two kinds of custom class, and S7 is right for only one of them.

| Kind | Classes | Built on | It is… |
| --- | --- | --- | --- |
| **Vector** types | `tm`, `fmls`, `mdl`, `mdl_tbl` | **vctrs** | a *vector of* terms / formulas / models |
| **Object** types | `mdl_gt`, `mdl_gt_build` | **S7** | one scalar specification or resolved build |

A `tm` is a vector: you can `c()` two of them, subset with `tm[1:3]`, cast it,
and drop it into a tibble column where **pillar** prints it. That behavior is
the whole point of the class, and it is *vctrs* machinery — records, proxies,
`vec_cast()`, `vec_ptype2()`, pillar shafts. S7 classes are **scalar objects
with typed properties**; they have none of that vectorization. Rewriting `tm`
as S7 would delete the very thing that makes it useful and force you to
rebuild it by hand, worse.

So the rule `recast` follows — the same one the tidyverse follows — is:

> **vctrs for vectors, S7 for objects.** They are complementary layers, not
> competitors. Use vctrs when the thing *is a vector*; reach for S7 when the
> thing is *one object* with named, typed parts.

The `<mdl_gt>` specification is squarely the second kind: one table's worth of
display instructions — selection, effect instructions, named cell groups,
layout, labels, and style. Before the migration it was a bare
`structure(list(...), class =
"mdl_gt")`. That is exactly the shape S7 improves on.

## Anatomy of an S7 class

Here is the spec's class definition (in `R/table-spec.R`), trimmed to its
skeleton:

```{r}
#| label: class-skeleton
#| eval: false
mdl_gt <- S7::new_class(
  "mdl_gt",
  properties = list(
    mdl_tbl   = S7_mdl_tbl,          # the fitted models (a registered vctrs type)
    family    = S7::class_data.frame,
    selection = S7::class_list,      # instruction slots the verbs fill
    labels    = S7::class_list,
    effects   = S7::class_list,
    groups    = S7::class_list,
    layout    = S7::class_list,
    style     = S7::class_list
  ),
  validator   = function(self) { ... },   # invariants, checked on every change
  constructor = function(object, ...) { ... }  # the gate
)
```

Three things are doing work here, and each earns its keep.

### 1. Properties are typed slots

Each `list()` slot became a **property** with a declared type. `S7::class_list`,
`S7::class_data.frame`, `S7::class_integer`, and `S7::class_any` are the
built-in type descriptors; a property typed this way rejects a value of the
wrong type at assignment. The type descriptors are ordinary values, so you can
name your own — which is how a *vctrs* type joins in (next section).

### 2. The validator runs on **every** modification — the headline win

This is the benefit that justified the whole migration. A validator returns
`NULL` (or `character(0)`) when the object is valid, or one string per
violation. Crucially, S7 runs it **on construction and again on every property
assignment**. `recast`'s spec validator is:

```{r}
#| label: validator
#| eval: false
validator = function(self) {
  msg <- character()
  preset <- self@layout$preset
  if (length(preset) &&
      !preset %in% c("adjustment", "levels", "interaction")) {
    msg <- c(msg, sprintf("@layout$preset must be ... not \"%s\"", preset))
  }
  d <- self@style$digits
  if (!is.null(d) && (!is.numeric(d) || length(d) != 1 || d < 0)) {
    msg <- c(msg, "@style$digits must be a single non-negative number or NULL")
  }
  # ... semantic axes, group ids, selection names ...
  msg
}
```

With the old bare list, the spec's invariants were re-checked by hand inside
each verb, and a bad *direct* assignment could sail through and only blow up at
render. Now the object cannot enter an invalid state at all:

```{r}
#| label: validator-demo
spec <-
  fmls(mpg ~ .x(wt) + hp, pattern = "sequential") |>
  fit(.fn = lm, data = mtcars, raw = FALSE) |>
  model_table(data = mtcars) |>
  mdl_gt()

# A nonsense layout is rejected the instant it is assigned
tryCatch(
  spec@layout$preset <- "banana",
  error = function(e) conditionMessage(e)
)

# So is a negative digit count
tryCatch(
  spec@style$digits <- -1,
  error = function(e) conditionMessage(e)
)
```

The verbs still validate their arguments for *friendly* messages; the validator
is the backstop that makes the guarantee total.

### 3. The constructor is the gate (and a sharp edge)

`mdl_gt(mt)` does real work before an object exists: keep only the fitted rows,
verify the table is one model family on one link and one presentable analysis,
stamp on the verified family table. That logic lives in a custom `constructor`.

One rule tripped the migration and is worth stating plainly:

> **S7 requires the literal `new_object()` call to appear in the constructor's
> own body.** It cannot be hidden inside a helper the constructor calls.

The first attempt delegated everything to a builder that ended in
`new_object()`, and S7 rejected it: *"`constructor` must contain a call to
`new_object()`."* The fix is to let the builder return the finished *property
values* and have the constructor make the `new_object()` call itself:

```{r}
#| label: constructor
#| eval: false
constructor = function(object, ...) {
  v <- build_mdl_gt(object, ...)   # all the gate logic; returns a named list
  new_object(
    S7_object(),
    mdl_tbl = v$mdl_tbl, family = v$family, selection = v$selection,
    labels = v$labels, columns = v$columns, layout = v$layout, style = v$style
  )
}
```

`new_object()`'s first argument is the parent instance (`S7_object()` for a
class with no S7 parent); the rest are the property values.

## Reading and writing properties: `@`, never `$`

S7 properties come off with `@`. On an S7 object, `$` is **not** property
access — and, gratifyingly, it does not silently return `NULL`; it errors:

```{r}
#| label: at-vs-dollar
spec@layout$preset            # `@` gets the property; then `$` indexes the list

tryCatch(spec$layout, error = function(e) conditionMessage(e))
```

That loud failure is a migration ally. Porting the codebase meant changing
`x$layout` to `x@layout` in a dozen files; every site that was missed failed
immediately at test time instead of hiding. (Note the two operators compose:
`spec@layout` gets the *list-valued property*, and `$preset` then indexes into
that plain list.)

Assignment goes back the same way and, as shown above, re-validates:

```{r}
#| label: at-assign
spec@layout$preset <- "levels"
spec@layout$preset
```

## Value semantics come for free

S7 objects are ordinary R values — copy-on-modify. The `recast` verbs
(`select_terms()`, `modify_style()`, …) each return a modified copy and never
mutate their input, which is exactly the lazy, composable contract the table
grammar wants:

```{r}
#| label: value-semantics
refined <- spec |>
  select_terms(~ wt) |>
  modify_style(digits = 4)

spec@style$digits      # the original spec is untouched (still unset)
refined@style$digits   # only the returned copy carries the instruction
```

## Generics and methods

`recast` uses all three ways S7 interacts with generics. Each is a small, real
example.

### An S7 generic of your own: `as_gt()`

`as_gt()` was an S3 generic (`UseMethod`); it became an S7 generic
(in `R/table-render.R`):

```{r}
#| label: s7-generic
#| eval: false
as_gt <- S7::new_generic("as_gt", "x")

method(as_gt, mdl_gt) <- function(x, ...) {
  built <- build_mdl_gt(x)
  render_cell_frame(built@cells, x)
}
```

`new_generic("as_gt", "x")` declares a generic dispatching on its first
argument; `method(generic, class) <- fn` registers a method.

### A method on a *base* generic: `print()` / `format()`

`print()` and `format()` are base S3 generics. The idiomatic S7 route is to
register methods on them rather than writing `print.mdl_gt`:

```{r}
#| label: base-method
#| eval: false
method(print,  mdl_gt) <- function(x, ...) { cat(format(x, ...), sep = "\n"); invisible(x) }
method(format, mdl_gt) <- function(x, ...) { ... }
```

```{r}
#| label: print-demo
spec   # dispatches to the S7 `format` + `print` methods
```

### An S7 method on someone else's *S3* generic: `adjustment_sets()`

`adjustment_sets()` is `recast`'s own S3 generic, and it must keep dispatching on
the vctrs `mdl_tbl`. To make the S7 `<mdl_gt>` join that same dispatch, register
an S7 *method* on the existing S3 generic:

```{r}
#| label: s7-on-s3
#| eval: false
method(adjustment_sets, mdl_gt) <- function(x, ...) {
  adjustment_sets(x@mdl_tbl, ...)
}
```

This works because S7 knows how to slot a method into S3 dispatch — but only
once you understand the namespacing wrinkle below.

## Interop: three things that will bite, and their fixes

Adopting S7 inside a package that already uses S3 and `inherits()` runs into a
handful of realities. `recast` hit all of them; here is the map.

### 1. In a package, an S7 class is **namespaced**

Define a class inside a package and its class string is `pkg::ClassName`, not
the bare name:

```{r}
#| label: namespacing
class(spec)
inherits(spec, "mdl_gt")        # FALSE! the class is "recast::mdl_gt"
inherits(spec, mdl_gt)          # TRUE — pass the class *object*, not a string
S7::S7_inherits(spec, mdl_gt)   # TRUE — the explicit S7 predicate
```

This happens whether or not you pass `package =` to `new_class()` — S7 infers
the package from where the class is defined. It has two consequences a mixed
codebase must handle:

- **`inherits(x, "mdl_gt")` string checks stop matching.** `recast` validates
  types through one helper, `validate_class()`, so the fix lived in one place:
  compare against the object's class names with any `pkg::` prefix stripped.

  ```{r}
  #| label: validate-class
  #| eval: false
  bareClasses <- sub("^[^:]*::", "", class(x))   # "recast::mdl_gt" -> "mdl_gt"
  if (!inherits(x, what) && !any(what %in% bareClasses)) stop(...)
  ```

- **S3 dispatch on the bare name stops finding the method.** A plain
  `adjustment_sets.mdl_gt` is never called, because dispatch sees
  `recast::mdl_gt`. That is *why* `adjustment_sets` needed the S7-method
  treatment above rather than a plain S3 method — S7 registers it under the
  namespaced class so dispatch resolves.

- **In tests**, `expect_s3_class(x, "mdl_gt")` fails for the same reason. Use
  the class object instead: `expect_true(S7::S7_inherits(x, mdl_gt))`.

### 2. Methods must be *registered* at load: `methods_register()`

S7 methods — including those on base and foreign S3 generics — are wired into
dispatch by a single call in `.onLoad()` (in `R/zzz.R`):

```{r}
#| label: onload
#| eval: false
.onLoad <- function(libname, pkgname) {
  # ...
  S7::methods_register()
}
```

Because of this, S7 methods need **no** `@export` and no `S3method()` entry in
`NAMESPACE`; only the *classes* and *generics* are exported. Omitting `@export`
on the `method(...)` calls (and letting `methods_register()` do the wiring) is
what keeps the `NAMESPACE` free of broken `S3method(print, mdl_gt)` lines
pointing at a `print.mdl_gt` that no longer exists.

### 3. Cross-file class references need `@include` (collation order)

`method(as_gt, mdl_gt) <-` in `table-render.R` mentions the `mdl_gt` *class
object*, which is created in `table-spec.R`. R sources files alphabetically by
default, and `table-render.R` sorts before `table-spec.R` — so at load the
class would not exist yet. An `@include` tag makes roxygen2 emit a `Collate:`
order that puts the definition first:

```{r}
#| label: include
#| eval: false
#' @include table-spec.R
#' @export
as_gt <- S7::new_generic("as_gt", "x")
```

Only files that reference a class **object** at load time need this; files that
merely mention `"mdl_gt"` as a *string* (e.g. `validate_class(x, "mdl_gt")`)
resolve at call time and need nothing.

## Registering a vctrs type so S7 can name it

The `mdl_tbl` property is typed by the vctrs `mdl_tbl` class. You register an S3
(or vctrs) class as an S7 type descriptor with `new_S3_class()` — *registering,
not converting*; `mdl_tbl` stays a full vctrs vector:

```{r}
#| label: register-vctrs
#| eval: false
S7_mdl_tbl <- S7::new_S3_class("mdl_tbl")
# ... then, in the class:  properties = list(mdl_tbl = S7_mdl_tbl, ...)
```

Now constructing a spec type-checks that what arrives really is a `mdl_tbl`, and
an S7 generic could dispatch on `mdl_tbl` the same way it dispatches on
`mdl_gt` — the two class systems referring to each other cleanly, each keeping
what it is good at.

## Two ends of the S7 spectrum

Not every S7 class needs the full apparatus. `recast`'s second S7 class,
`mdl_gt_build` (the resolved output of the build orchestrator), is a *plain
value* — all typed properties, no behavior:

```{r}
#| label: build-class
#| eval: false
mdl_gt_build <- S7::new_class(
  "mdl_gt_build",
  properties = list(
    selected   = S7::class_data.frame,
    effects    = S7::class_data.frame,
    conditions = S7::class_data.frame,
    measures   = S7::class_data.frame,
    groups     = S7::class_data.frame,
    layout     = S7::class_list,
    cells      = S7::class_data.frame
  )
)
```

No custom constructor (S7 generates one from the properties), no validator (no
cross-field invariant to guard). It earns its keep simply by letting the
resolver return one **self-documenting, type-checked** object instead of a bare
list. Put next to `mdl_gt`, it shows the range: reach for a validator and a
custom constructor only when there is a gate to guard or an invariant to keep.

## Adoption checklist

Everything above, as a sequence for the next class:

1. **Decide it's an object, not a vector.** If you want `c()`, `[i]`, casting,
   or pillar columns, it is a vctrs type — stop here.
2. `Imports: S7` in `DESCRIPTION`; `S7::methods_register()` in `.onLoad()`.
3. `new_class()` with typed `properties`; add a `validator` for real
   invariants and a `constructor` only if construction is more than filling
   slots (and put `new_object()` in the constructor's own body).
4. Register any vctrs/S3 types you reference with `new_S3_class()`.
5. Generics with `new_generic()`; methods with `method(gen, cls) <- fn` —
   including on base and foreign S3 generics. Export classes and generics;
   **not** methods.
6. Add `@include` wherever one file names another's class object at load.
7. Read properties with `@`; expect a namespaced `pkg::Class`, so test with
   `S7_inherits()` and make any `inherits(x, "name")` string checks
   prefix-tolerant.

The reward: typed slots, invariants enforced on every change, clean dispatch,
and value semantics — an *object* layer that composes naturally on top of the
*vector* layer vctrs already gives you.
```
