---
title: "Internals: how the machine is built"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Developer-facing companion to `vignette("usage")`. Shows the data structure
under each layer and the seams between them, using `recast:::` internals where
needed. Each section ends with **review questions** — the places where the
design deserves a second pair of eyes. Concrete defects live in `issues.md`.

```{r setup}
library(recast)
d <- mtcars
d$cyl <- factor(d$cyl)
```

## 1. A term is a record

`tm` is a `vctrs` record: parallel fields, one slot per term. The proxy *is*
the term table — every later structure is built from this.

```{r}
t <- tm(mpg ~ .x(wt) + .s(am) + log(hp))
vctrs::vec_proxy(t)
```

`tm.formula()` is a fixed pipeline over the syntax tree:

1. `parse_formula_terms()` — recursive descent; runes, `a:b`, `(1|g)`,
   transformations, opaque calls (`Surv(...)`) each handled at the node
2. `demote_orphan_roles()` — `.i()`/`.m()` without an exposure demote to
   predictors, with a warning
3. `expand_shortcut_interactions()` — each `.i()` crosses each exposure into
   an explicit `x:i` term, group-tiered so the pair travels together
4. `apply_default_roles()` — position fills in what wasn't declared
5. `apply_term_arguments()` — explicit `role = `/`label = ` arguments override

The raw parse, before defaults:

```{r}
recast:::parse_formula_terms(mpg ~ .x(wt) + .i(cyl))
```

Note `side`: strata and random effects get `side = "meta"` — neither left nor
right. That single value drives everything in §3.

**Review questions**

- `log(hp)` keeps its full call as its *name* and records the wrapper in
  `transformation`. Two representations of one fact — is the field earning
  its keep, or should the call be parsed once at `set_data()`?
- Roles live in a flat character field; `group` is a separate integer tier.
  Interactions need both to travel together. Simpler as one composite?

## 2. A `fmls` is a membership matrix plus a term table

Two structures, stitched by term name:

```{r}
f <- fmls(mpg ~ .x(wt) + hp, pattern = "sequential")
vctrs::vec_data(f)      # rows = formulas, cols = terms, 1 = membership
attr(f, "termTable")    # scalar attribute: one definition per term
```

Patterns are functions `tm -> tbl_df` (a registry, `register_pattern()`);
their output is a *precursor table*, one row per future formula:

```{r}
apply_sequential_pattern(tm(mpg ~ .x(wt) + hp))
```

`fmls()` then flattens each precursor row into the 0/1 matrix directly —
each row's unique non-`NA` terms become that row's memberships, so a term
repeated within one precursor row still records `1`.

Every registered pattern honors that contract now,
`apply_fundamental_pattern()` included: the exposure keeps its key-pair
column and each remaining term rides one row as `covariate_1`. Because the
fundamental pattern reduces each formula to a single left- and right-hand
term, `fmls(pattern = "fundamental")` demotes meta terms (strata, random
effects) to plain predictors before expansion, with a message:

```{r}
fmls(mpg ~ wt + .s(am), pattern = "fundamental")
```

**Review questions** *(both resolved 2026-07-09)*

- ~~The seven-role extraction block is copy-pasted through every pattern.~~
  `pattern_roles()` pulls every role once and `key_pair_grid()` builds the
  outcome × exposure grid; the patterns, `check_mediation()`, and
  `check_groups()` all draw from them.
- ~~The `table()`-based matrix build records a count of 2 for a repeated
  term.~~ The membership build above is 0/1 by construction, and the
  downstream tests read `>= 1`.

## 3. Where the strata goes (and why formulas drop it)

`formula.tm()` rebuilds from `side == "left"` and `side == "right"` only.
Random effects are re-added as `(1 | g)`; **strata is not re-added anywhere**:

```{r}
f <- fmls(mpg ~ .x(wt) + .s(am))
formula(f)
```

The stratum re-enters at `plan_fit()` — not as formula syntax but as *data
splitting*: one plan row per stratum level, the level's rows subset before
the engine sees anything.

```{r}
plan_fit(f, data = d)
```

Meta terms are members of the formula matrix like any other term, and
`formulas_to_terms()` reads membership alone — so a stratum applies only to
the family that declared it, and combining families cannot leak it:

```{r}
combined <- suppressMessages(c(f, fmls(qsec ~ .x(hp))))
plan_fit(combined, data = d)[, c("formula_call", "strata_variable")]
```

A stratifying term must actually exist in the fitting data: a missing
column would enumerate zero levels, and the crossing would erase every
model of that formula from the plan without a trace (`fit()` returned
`<model[0]>` silently). Since 2026-07-10 `plan_fit()` errors on the
mismatch instead:

```{r}
#| error: true
f |> add_strata(not_a_column) |> plan_fit(data = d)
```

Family identity works the same way as the stratum: *derived from roles,
not recorded as lineage*. A `fmls` is born as one family — one expansion
pattern generated its rows — but `c()` just stacks rows and merges term
tables; nothing marks where each row came from. The internal
`identify_families()` is the role-reading recovery: mediators bind their
triads back together, and adjustment-ladder signatures relate families across
exposures or outcomes. A `mdl_tbl` carries the result as its `family`,
`pattern`, and `relation` columns, restamped whenever the table is built or
reshaped, so no user call is needed. Everything downstream re-derives from the
same roles — the `mdl_gt()` gate verifies one analysis, and
`adjustment_set_index()` keys each model's rung by its actual covariate set
(see the usage vignette). Carrying a family id through `fit()` as recorded
lineage remains the open item in `issues.md`.

**Review questions**

- ~~The stratum is invisible everywhere models print.~~ *Resolved
  2026-07-09*: the printed family annotates meta terms in rune form
  (`.s(am)`, `.r(id)`) after the right-hand terms, while `formula()`,
  `formula_call`, and the recorded call stay the true fitted formula —
  they are consumed programmatically (re-fitting, term counting), so
  annotating them would corrupt real syntax. The `strata`/`level` columns
  remain the machine-readable context.
- ~~How do engine-native strata (`survival::strata()` in a coxph formula)
  coexist with `.s()`?~~ *Decided 2026-07-09*: they coexist as the two
  different mechanisms they are. `survival::strata()` *conditions within*
  one model — the engine understands it, so the call passes through the
  formula untouched (whole, as one term), rideable like any covariate and
  traced as `transformation = "strata"`. `.s()` remains the grammar's own
  construct: an actual data split, one fit per stratum level. Neither is
  rewritten into the other:

  ```{r}
  t <- tm(Surv(time, status) ~ age + strata(sex))
  vctrs::vec_proxy(t)[, c("term", "side", "role", "transformation")]
  ```

## 4. Why data attaches to the *table*, not the formula

Formulas stay abstract on purpose — a `fmls` captures no data, so families
are cheap to build, combine, and store. But three consumers need real data
at realization time:

- **categorical levels** — the term table's `level` field is empty until
  `set_data()`; the table layer stamps it from attached data to know that
  `cyl8` is level `8` of `cyl` (`resolve_term_metadata()`)
- **events and rates** — `add_events()` runs `survival::pyears()` on the
  attached data at render
- **interaction rows** — per-level `n` counts come from the attached data

So the dataset attaches once, to the `mdl_tbl`, keyed by name:

```{r}
mt <-
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential") |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

names(attr(mt, "dataList"))
term_table(mt)
```

**Review questions**

- ~~The link is a fragile *name* match.~~ *Resolved 2026-07-09*: a `data`
  passed as an inline expression takes a stable content-derived id
  (`data_<hash>`) at both `fit()` and `attach_data()`, so identical content
  meets itself; and a frame arriving under a different name is aliased —
  with a message — to the one referenced-but-detached `data_id` whose
  models' variables it fully carries (the aliasing logic inside
  `attach_data()`). An explicit
  `name = ` always wins.
- ~~Attaching whole data frames makes a `mdl_tbl` heavy to serialize.~~
  *Decided 2026-07-09*: the frame attaches whole, deliberately — pruning to
  the referenced columns was tried and walked back, because later work
  routinely reaches for columns no current formula names (an `add_events()`
  follow-up column, a variable for the next family of models added to the
  same table). What the decision *did* pin down is the layering:
  `set_data()` on a `tm` or `fmls` only *teaches* — it stamps term
  properties and retains nothing, because formulas stay abstract and the
  source data keeps evolving — while the `mdl_tbl`, where formulas and data
  come together, is the one layer that retains data.

## 5. A `mdl` is a deconstructed model

Not the fitted object — a blueprint: call name, `fmls`, tidy estimates,
glance summary, data context. Enough to present and diagnose, cheap to keep.

```{r}
m <- fit(fmls(mpg ~ .x(wt) + hp), .fn = lm, data = d)
vctrs::field(m, "parameterEstimates")[[1]]
vctrs::field(m, "dataArgs")[[1]]
```

Notes:

- `execute_fit()` rewrites the recorded call (function by name, data by
  name) so models hash and compare cleanly.
- Failures are soft: an error becomes a `mdl` whose `summary_info$error`
  carries the message; `fit_status` turns `FALSE` downstream.
- The `var_cov` matrix is stored per model — that is what lets
  `estimate_interaction()` derive within-level effects without refitting.

**Review questions** *(both resolved 2026-07-09)*

- ~~`mdl.lm` and `mdl.lmerMod` are ~80% identical.~~ Both are thin
  wrappers now: each computes its own tidy/glance pieces and hands the
  rest to `new_model_from_fit()`, the shared trunk.
- ~~`my_tidy()` exposes an `exponentiate` argument it ignores.~~ The
  parameter is gone; exponentiation stays deferred to `flatten_models()`.

## 6. The `mdl_tbl`: invariant columns, reconciled attributes

A tibble subclass — one row per model — carrying three scalar attributes
that must stay parallel to the rows: the formula matrix, the term table, the
data list. Every subset/filter reconciles them (`df_reconstruct()`):

```{r}
sub <- dplyr::filter(mt, number < 3)
class(sub)
formula_matrix(sub)      # trimmed to the surviving rows and their terms
```

Dropping an invariant column deliberately demotes the object:

```{r}
class(dplyr::select(mt, -fit_status))
```

**Review questions**

- ~~`formula_index` stores the membership row itself, and nothing consumes
  it.~~ *Resolved 2026-07-09*: dropped. The formula matrix's rows stay
  parallel to the table's, so a row-index column carried no information.
- Rows link to attributes by `id = rlang::hash(model)`. Two identical fits
  (same formula, same data) collide by design — is dedup ever wrong?

## 7. The `mdl_gt`: five stages from spec to `{gt}`

`mdl_gt()` itself is the gate between the notebook and the table: it verifies
the fitted rows form *one presentable analysis* — a single formula family, or
families bound by a shared `relation` over *one* adjustment
ladder — and records the verified structure on the spec (`spec$family`).
Narrowing by outcome, exposure, or strata happens before the gate, on the
`mdl_tbl` (the paring verbs — `keep_outcomes()`, `restrict_to()`,
`adjusting_for()`, `excluding()`, ... — or plain `dplyr`); the spec's own
selection holds only display decisions. Within the resolver, an adjustment set's index is
the *identity* of its covariate set (`adjustment_set_index()`, displayed by
`adjustment_sets()`), not its position within a family — which is what lets
related families' rows align on one mesa.

A `<mdl_gt>` is an S7 specification of *instructions* — nothing is computed
until inspection or render:

```{r}
spec <- mt |> mdl_gt() |> add_estimates()
spec@selection
spec@groups
spec@layout
```

Runs in fixed stages, each file owning one:

| stage | representation | file |
|---|---|---|
| specify | S7 `mdl_gt` + public verbs | table-spec.R |
| build | effects, conditions, atomic measures, groups | table-build.R |
| layout | semantic axes → explicit cell frame | table-layout.R |
| render | cell frame → `{gt}` | table-render.R |

The effect frame — one row per effect, with categorical reference anchors:

```{r}
inspect_mdl_gt(spec, "effects")[, c("outcome", "adjustment", "term",
                                     "contrast", "estimate", "is_reference")]
```

The **cell frame** — the single contract between layout and render, one row
per rendered cell; `render_cell_frame()` is the only place `{gt}` calls are
emitted:

```{r}
inspect_mdl_gt(spec, "cells")
```

Two mechanisms live only in the renderer:

- **Group-scoped cells** (the interaction p-value floating over its level
  rows): `{gt}` has no body rowspan, so the value is written into every row
  and all but one copy is *blanked by a text transform*
  (`apply_group_scoped()`) — a content substitution, which holds on dark
  themes and every output format (the old white, zero-size text styling
  did not).
- **Plot cells** (the forest column): each cell is a ggplot drawn as vector
  SVG and inlined as its own base64 `data:` document, sized in `em` units
  so it scales with the text beside it (`plot_image()`, which pins both
  dimensions — the svg device rounds its canvas to whole points, and the
  rounding would otherwise skew the axis strip's width off the cells'; a
  build without cairo falls back to the fixed-size PNG). The column's
  native look is a *whole-body* default, not per-cell styling: with a plot
  column present the body renders borderless (the journal booktabs
  convention), because under `border-collapse: collapse` a per-cell hidden
  border wins every shared edge and punched gaps in the header and bottom
  rules. The axis strip continues the cells' dashed reference line and
  takes an optional `title` beneath its tick labels.

The unit everything above begins from is an **effect**: outcome × focal
term/contrast × model context, optionally conditioned by strata and modifier
levels. Atomic measures declare their own grains. Named presentation groups
then read those measures, and a semantic layout projects them into explicit
row and column ids. Identity is carried by fields and stable ids; no renderer
recovers it by parsing a display key.

**Review questions**

- ~~Can presentation groups move without changing statistics?~~ *Implemented
  2026-07-14*: `place_cells()` moves stable groups among columns, statistic
  rows, and the body; `modify_layout()` moves semantic dimensions; both consume
  unchanged effects and measures.
- ~~`frame_context()` is called with `dec = NULL` on the interaction
  path.~~ *Resolved 2026-07-09*: the unused parameter is gone;
  `frame_context(spec)` takes the specification alone.

## 8. Review agenda

Ordered start-points, cross-referenced to `issues.md`. The 2026-07-09
repair pass closed the outstanding items: the high-severity contract fixes
(the fundamental-pattern contract, the strata leak on family combination,
the two definitions of "categorical", the interaction layout and estimator),
the medium design-friction items (`fit(raw = )` now defaults to `FALSE`,
strata annotate in printed formulas, forest cells draw as scalable SVG, the
group mask survives themes and formats, the one-family check compares
links, data attaches by content when names diverge), and the parsimony
passes (§2, §5, the sequential pattern, `rhs()`'s tree split).

The three design questions that remained open were decided on 2026-07-09:

1. Engine-native `strata()` is *conditioning*, not a data split: it passes
   through the formula untouched (traced in `transformation`), while `.s()`
   remains the grammar's own data-splitting stratum (§3)
2. The effect is the semantic unit; measures declare grain, named cell groups
   declare presentations, and axes/placements remain independent (§7)
3. Attached data attaches whole, at the `mdl_tbl` level only —
   `set_data()` teaches, the `mdl_tbl` retains (§4)
