---
title: "Model-table gallery"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(recast)
d <- mtcars
d$cyl <- factor(d$cyl)
d$am <- factor(d$am)
```

This gallery is the visual acceptance surface for the effect-and-presentation
grammar. Every example travels through the same semantic build and layout
compiler; only effects, axes, cell-group placement, and style differ.

## Multi-outcome adjusted regression

```{r}
multiOutcome <- c(
  fmls(mpg ~ .x(cyl) + hp, pattern = "sequential"),
  fmls(qsec ~ .x(cyl) + hp, pattern = "sequential")
) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

multiOutcome |>
  mdl_gt() |>
  add_n() |>
  add_estimates() |>
  modify_style(theme = "journal") |>
  as_gt()
```

## Survival table with event/rate rows and rate difference

```{r}
#| eval: !expr requireNamespace("survival", quietly = TRUE)
lung <- survival::lung
lung$sex <- factor(lung$sex, levels = 1:2, labels = c("Male", "Female"))
survivalModels <-
  fmls(Surv(time, status) ~ .x(sex) + age, pattern = "sequential") |>
  fit(.fn = survival::coxph, data = lung) |>
  model_table(data = lung)

survivalModels |>
  mdl_gt() |>
  modify_layout(preset = "levels") |>
  add_events() |>
  add_rate_difference() |>
  add_estimates(columns = list(beta ~ "HR", conf ~ "95% CI")) |>
  as_gt()
```

## Interaction forest table

```{r}
interactionModels <-
  fmls(mpg ~ .x(hp) + .i(cyl)) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

interactionModels |>
  mdl_gt() |>
  add_interaction() |>
  add_n() |>
  add_estimates() |>
  add_forest(axis = list(title = "Mean difference")) |>
  as_gt()
```

## Combined stratum × modifier table

```{r}
conditionedModels <-
  fmls(mpg ~ .x(hp) + .i(cyl) + .s(am)) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

conditionedModels |>
  mdl_gt() |>
  add_interaction() |>
  add_estimates() |>
  add_forest() |>
  modify_layout(
    rows = c("stratum", "stratum_level", "modifier", "modifier_level"),
    columns = c("term", "contrast")
  ) |>
  as_gt()
```

## Wide categorical table with nested spanners

```{r}
wideCategorical <- c(
  fmls(mpg ~ .x(cyl) + hp, pattern = "sequential"),
  fmls(mpg ~ .x(am) + hp, pattern = "sequential")
) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

wideCategorical |>
  mdl_gt() |>
  add_estimates(view = "separate") |>
  modify_layout(
    rows = c("outcome", "adjustment"),
    columns = c("term", "contrast")
  ) |>
  as_gt()
```
