---
title: "Usage: terms → formulas → models → tables"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Working surface for interactive development. One continuous walkthrough of the
pipeline, layer by layer — point at a chunk, say what looks wrong, we repair
it. Succinct by intent; the machinery behind each chunk is in
`vignette("internals")`.

```{r setup}
library(recast)
d <- mtcars
d$cyl <- factor(d$cyl)
d$am <- factor(d$am, levels = c(0, 1), labels = c("Automatic", "Manual"))
```

## 1. Terms

Roles declared inline with runes (`.x()` exposure, `.o()` outcome, `.s()`
strata, `.i()` interaction, `.m()` mediator, ...):

```{r}
t <- tm(mpg ~ .x(wt) + hp + .s(am))
t
describe(t, "role")
```

Terms become data-aware when they meet a dataset — type, distribution, and
levels get stamped on:

```{r}
t <- set_data(t, d)
describe(t, "level")
```

## 2. Formulas

A `fmls` is a *family*: one recipe, many formulas, expanded by a pattern.

```{r}
fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential")
```

```{r}
fmls(mpg ~ .x(wt) + hp + cyl, pattern = "parallel")
```

The listing *is* the display: roles show through the term coloring rather
than a legend, and `describe()` reads them off a family directly:

```{r}
describe(fmls(mpg ~ .x(wt) + hp), "role")
```

Strata ride along as *meta* terms — annotated in the printed family with
their declared rune, absent from the fitted formula itself:

```{r}
f <- fmls(mpg ~ .x(wt) + hp + .s(am))
f
formula(f)
```

> ✓ **Resolved (2026-07-09).** The stratum splits data at `fit()` — it is not
> a covariate — so `formula()` and the `formula_call` column stay the true
> fitted (unstratified) formula. The *printed* family now carries `.s(am)`
> after the right-hand terms (random effects annotate as `.r(...)` the same
> way), so a stratified model can no longer be misread as unstratified.

Engine-native stratification is a *different* mechanism, and stays one:
`survival::strata(...)` conditions within a single model, so it passes
through the formula untouched (traced as `transformation = "strata"`),
while `.s()` splits the data into separate fits:

```{r}
fmls(Surv(time, status) ~ .x(age) + strata(sex))   # conditioning: one model
fmls(Surv(time, status) ~ .x(age) + .s(sex))       # data split: one per level
```

Interactions cross with the exposure automatically:

```{r}
fmls(mpg ~ .x(hp) + .i(cyl))
```

A mediator triggers the causal triad:

```{r}
fmls(mpg ~ .x(wt) + .m(hp) + qsec)
```

Families are playable after the fact:

```{r}
fmls(mpg ~ .x(wt) + hp) |>
  add_strata(am) |>
  subset_data(disp > 100)
```

And whether formulas are *related* — one family or several, and how — is
readable from the causal roles themselves. A `model_table()` records this
automatically, sorting its rows into families and carrying a `family`,
`pattern`, and `relation` column that name each family's pattern and the
relation between families that share an adjustment ladder (no separate call,
and refreshed whenever the table is reshaped):

```{r}
c(
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential"),
  fmls(mpg ~ .x(disp) + hp + cyl, pattern = "sequential")
) |>
  model_table()
```

Same outcome, same ladder, different exposures: the `varied exposures`
relation supplies a semantic exposure/term dimension over shared adjustment
rows. A mediation triad binds into a single family even though
its `mediator ~ exposure` member has a different left-hand side:

```{r}
fmls(mpg ~ .x(wt) + .m(hp) + qsec) |>
  model_table()
```

## 3. Models

The plan comes before the fit — formula × stratum × subset, inspectable:

```{r}
fs <- fmls(mpg ~ .x(wt) + hp, pattern = "sequential") |> add_strata(am)
plan_fit(fs, data = d)
```

`fit()` keeps the causal context attached by default (a `mdl` vector;
`raw = TRUE` opts out for a quick look at the plain fits); `model_table()`
is the notebook of models:

```{r}
m <- fit(fs, .fn = lm, data = d)
mt <- model_table(m, data = d)
mt
```

> ✓ **Resolved (2026-07-09).** `raw = FALSE` is now the default, so the main
> path no longer repeats the argument at every call.

Attached data is retained at the `mdl_tbl` level only, and the frame
attaches whole — later work routinely reaches for columns no current formula
names (a follow-up column, a variable for the next family). `set_data()` on
terms or formulas *teaches* term properties and retains nothing; the model
table is where formulas and data actually meet.

```{r}
summary(mt)
```

Estimates, one row per parameter:

```{r}
flatten_models(mt)
```

## 4. Tables — the mesa

```{r}
mt2 <-
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential") |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)
```

### The handoff: pare the table, verify, then lay out

*Which models* sit on a mesa is the `mdl_tbl`'s decision, made with ordinary
`dplyr` verbs before `mdl_gt()` is ever called. The table already carries its
family structure — `family`, `pattern`, `relation`, refreshed on every
reshape — so the paring can go by analysis rather than by column value, and
the print shows what would sit together:

```{r}
mtMixed <-
  c(
    fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential"),
    fmls(mpg ~ .x(disp) + hp + cyl, pattern = "sequential"),
    fmls(qsec ~ .x(am))
  ) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)
mtMixed
```

`mdl_gt()` is then the gate: it verifies what arrives is *one presentable
analysis* — a single family, or families bound by a shared relation over one
adjustment ladder (varied exposures / varied outcomes, the wide-table
shapes) — and turns anything looser back toward the table's own verbs:

```{r}
#| error: true
mtMixed |> mdl_gt()
```

The paring itself speaks the causal language, one verb per dimension, so
a pipe reads like the analysis sentence. `keep_outcomes()` /
`keep_exposures()` (and their `drop_*()` complements) go by causal role;
`keep_families()` by the identified structure; `restrict_to()` by the
population (stratum, level, subset rule) or dataset; `adjusting_for()`
keeps the models whose adjustment set carries the named covariates; and
`excluding()` keeps the models whose formulas avoid a term entirely — the
way to set aside a mediator before laying the rest out. Everything is
matched exactly, a typo errors with what is available instead of silently
keeping nothing, and each verb reports what it kept. Plain `dplyr::filter()`
still works; these are the shorthand whose result is likelier to pass the
gate:

```{r}
mtMixed |> keep_outcomes(mpg) |> mdl_gt()
```

```{r}
# Family ids come from the `family` column printed above — look first, then cut
mtMixed |> keep_families(1, 2) |> mdl_gt()
```

```{r}
# Or by what the models adjust for
mtMixed |> adjusting_for(hp, cyl)
```

Past the gate, every verb is a *display* decision: which adjustment rungs
show (`select_adjustment()`), which terms' cells show (`select_terms()`),
what the labels say (`modify_labels()` — outcomes included), and which named
cell groups are active (`add_*()`). `modify_layout()` maps semantic dimensions
and `place_cells()` moves those groups without changing measures. The rungs
need not be remembered —
`adjustment_sets()` shows each distinct covariate set, what it adds over the
rung below, and the number `select_adjustment()` picks it by. The index is
the *identity* of the set (not its position within a family), so related
families sharing a ladder share rung numbers and their rows align on the
mesa by the actual adjustment:

```{r}
adjustment_sets(mtMixed)
```

A bare `mdl_gt()` is already a table; printing shows the specification without
rendering:

```{r}
mt2 |> mdl_gt()
mt2 |> mdl_gt() |> as_gt()
```

Grown one verb at a time — select, add, modify, in any order:

```{r}
mt2 |>
  mdl_gt() |>
  select_adjustment(1 ~ "Unadjusted", 3 ~ "Fully adjusted") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI")) |>
  modify_labels(wt ~ "Weight (1000 lbs)") |>
  as_gt()
```

### The `"levels"` layout: events, rates, hazard ratios

```{r}
#| eval: !expr requireNamespace("survival", quietly = TRUE)
lung <- survival::lung
lung$sex <- factor(lung$sex, levels = 1:2, labels = c("Male", "Female"))

mtSurv <-
  fmls(Surv(time, status) ~ .x(sex) + age, pattern = "sequential") |>
  fit(.fn = survival::coxph, data = lung) |>
  model_table(data = lung)

mtSurv |>
  mdl_gt() |>
  modify_layout(preset = "levels") |>
  select_adjustment(1 ~ "Unadjusted", 2 ~ "Age-adjusted") |>
  add_events(followup = time) |>
  add_rate_difference() |>
  add_estimates(columns = list(beta ~ "HR", conf ~ "95% CI")) |>
  as_gt()
```

### The `"interaction"` layout, with a forest column

```{r}
mtInt <-
  fmls(mpg ~ .x(hp) + .i(cyl)) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

mtInt |>
  mdl_gt() |>
  add_interaction() |>
  add_n(label = "No.") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI",
                               p ~ "P for interaction")) |>
  add_forest(axis = list(title = "Estimate")) |>
  as_gt()
```

> ✓ **Resolved (2026-07-09).** Forest cells now draw as inline vector SVG
> sized in `em` units, so they scale with the text they sit beside (a build
> without cairo falls back to the old fixed-size PNG). See
> `vignette("internals")` §7.
>
> ✓ **Resolved (2026-07-10).** The forest column now reads as native to the
> table: with a plot column present the whole body renders borderless (the
> journal booktabs look) instead of the plot cells hiding their own borders
> — which, under CSS border collapsing, punched gaps in the header and
> bottom rules; the axis strip is pinned to exactly the cells' width (the
> svg device's whole-point rounding used to skew it); the dashed reference
> line runs down into the axis; and `add_forest(axis = list(title = ))`
> draws an axis title beneath the tick labels, as above.

### Style

```{r}
mt2 |>
  mdl_gt() |>
  add_estimates() |>
  modify_style(accents = list(p < 0.05 ~ "bold"), digits = 3) |>
  as_gt()
```
